---
title: "An overview of Difference-in-Difference and Synthetic Control Methods: Classical and Novel Approaches"
description: |
  Society for Epidemiologic Research (SER) Workshop:   |  
  Setup: Part 1/4
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Roch Nianogo
    url: https://ph.ucla.edu/about/faculty-staff-directory/roch-nianogo
    affiliation: Department of Epidemiology, UCLA Fielding School of Public Health
    affiliation_url: https://ph.ucla.edu/about/faculty-staff-directory/roch-nianogo
date: "April 22th 2024"
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---


```{r, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 400
)
```






In this document, we will provide all steps and R codes required to implement 
the methods outlined in the slides. This will include the following methods

- Interrupted Time Series (ITS) 
- Difference-in-difference (DID) and related methods 
- Synthetic control methods and related methods   

There will be four html files  
- The setup (this file) (1/4)   
- Difference-in-difference (DID) and related methods (2/4)   
- Synthetic control methods and related methods (3/4)  
- Heterogeneous treatment and staggered adoption (4/4)  





**Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at niaroch@ucla.edu or tbenmarhnia@ucsd.edu**.



# Loading packages  

To reproduce the html documents, you will need to install the following 

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R,
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides 
the template for this document.

Once everything is set up, we load the following packages:

```{r}

if (!require("pacman")){
  install.packages("pacman", repos = 'http://cran.us.r-project.org')
} # a nice package to load several packages simultaneously


p_load("tidyverse","magrittr","broom",        #for manipulating data
       "cleaR",                               #for clearing workspace
       "here",                                #for directory managment
       "Synth",                               #for the traditional synthetic control method
       "gsynth",                              #for the generalized synthetic control method
       "panelView",                           #for presenting data in a panel
       "lme4",                                #for multi-level analysis
       "estimatr",                            #for robust linear model
       "did",                                 #for staggered difference-in-difference
       "gtsummary")                           #for nice tables
```

# Loading the data  

This is a simulated data where a policy (e.g. smoking ban) was implemented   
- The policy was enacted in five states at the same time: Alabama,  Alaska, Arizona, Arkansas and California   
- The policy was enacted in **2000**  
- The unit of analysis is the **state**  
- `y`  is the outcome  
- `xi` is a time-invariant variable (but varies across states)  
- `xt` is a time-varying variable (but is constant across states)  
- `xit` is a time-varying and unit-variable (i.e. varies by year and state)    

Let's load the data           

```{r load_data}
mydata <- read_csv(here("data", "sim_data.csv"))
```

Create some useful variables within the datasets:  
- an indicator for after the policy has been implemented: called this `post`  
- an indicator for states that have implemented the policy: called this `treated`  
- an interaction term between post and treated: called this `treatedpost`  
- create a recentered year variable such that `year_rec = 0` at the time of the policy  



```{r edit_data}
year_policy <- 2000

mydata <- mydata %>% 
  mutate(year_rec = year - year_policy,
         post     = ifelse(year>=year_policy,1,0),
         treated  = ifelse(state %in% c("Alabama",  "Alaska", 
                                        "Arizona", "Arkansas", "California"), 1,0),
         treatedpost = treated*post)
```
# Exploring the data structure  


```{r exploring_data, layout="l-body-outset"}
library(rmarkdown)
paged_table(head(mydata, 10))

glimpse(mydata)

p_load("skimr")
skim(mydata)
```


# Visualizing the data and policy  

## Treatment overview  
```{r visualizing}
mydata <- mydata %>% 
  mutate(year_rec = as.integer(year_rec)) %>%
  as.data.frame() # need to convert to a data.frame for some functions to work
```

Using PanelView    

```{r}
p_load("panelView")
panelview(y ~ treatedpost, data = mydata,
          index = c("state","year_rec"), 
          pre.post = TRUE) 
```

Using Panel Match   

```{r}
p_load("PanelMatch")
DisplayTreatment(unit.id = "state_num",
                 time.id = "year_rec", legend.position = "none",
                 xlab = "year", ylab = "state",
                 treatment = "treatedpost", data = mydata)
```



## Sample tables and Covariate balance plots      
### Table 1    
```{r tables0}
p_load("gtsummary")
n_treated0 <- mydata %>% 
  filter(treated == 0) %>% 
  select(state) %>% 
  n_distinct()

n_treated1 <- mydata %>% 
  filter(treated == 1) %>% 
  select(state) %>% 
  n_distinct()

tab1 <- mydata %>%
  filter(post==0) %>% 
  select(c("xit","xt", "xi", "treated","y")) %>% 
  mutate(treated= case_when(treated==1~"Treated",
                            TRUE~"Control")) %>% 
  tbl_summary(
    missing = "no",
    by ="treated",
    type = list(everything() ~ "continuous"),
    digits = list(everything() ~ 2),
    statistic = list(everything()~"{mean}")
  ) %>% 
  modify_spanning_header(c("stat_1", "stat_2") ~ "Table 1") %>% 
  modify_header(update = list(
    label ~ '**Characteristic**',
    stat_1 ~ '**Control**, N = {n_treated0}',
    stat_2 ~ '**Treated**, N = {n_treated1}'
  )) 
tab1
```

### Covariate balance plots  
Balance by treated group    

```{r}
p1 <- mydata %>%
  filter(post==0) %>% 
  select(xt, xi, xit, y, treated) %>% 
  mutate(treated= case_when(treated==1~"Treated",
                            TRUE~"Control")) %>% 
  group_by(treated) %>% 
  group_modify(~ {.x %>% map_dfr(mean)}) %>% 
  pivot_longer(cols = -treated,
               names_to = c("variable"),
               values_to = "mean") %>% 
  ggplot(aes(x=treated, y=mean, fill=factor(treated))) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label = round(mean,2)), 
            position = position_stack(0.5), 
            size=3, 
            color = "black")+
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Checking for imbalance in variables pre-policy",
       y = "Mean",
       x = "Variables",
       fill = "Treatment status")+
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
p1
```
Notice xt is constant across units   


Balance by post period   

```{r}
p2 <- mydata %>%
  filter(treated==0) %>% 
  select(xt, xi, xit, y, post) %>% 
  mutate(post= case_when(post==1~"After",
                         TRUE~"Before")) %>% 
  group_by(post) %>% 
  group_modify(~ {.x %>% map_dfr(mean)}) %>% 
  pivot_longer(cols = -post,
               names_to = c("variable"),
               values_to = "mean") %>% 
  ggplot(aes(x=post, y=mean, fill=factor(post))) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label = round(mean,2)), 
            position = position_stack(0.5), 
            size=3, 
            color = "black")+
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Checking for imbalance in variables in control units",
       y = "Mean",
       x = "Variables",
       fill = "Time")+
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
p2
```
Notice xi is constant across time  


## Time trend plots  
For all the data   

```{r tables}
mydata %>% 
  ggplot(aes(x=year, y=y, group=state)) + 
  annotate("rect", fill = "gray", alpha = 0.5,
           xmin = 2000, xmax = 2010,
           ymin = -Inf, ymax = Inf) +
  labs(title = paste("Outcome by year"),
       x = "Year", 
       y = "Outcome",
       color = "Treatment") +
  geom_line(aes(color=factor(treated)), size=0.5) +
  scale_color_discrete(labels=c("Control", "Treated")) +
  geom_vline(xintercept = year_policy, lty=2) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

For two states   

```{r}
mydata %>% 
  filter(state %in% c("California", "Georgia")) %>% 
  ggplot(aes(x=year, y=y, group=state)) + 
  annotate("rect", fill = "gray", alpha = 0.5,
           xmin = 2000, xmax = 2010,
           ymin = -Inf, ymax = Inf) +
  labs(title = paste("Outcome by year"),
       x = "Year", 
       y = "Outcome",
       color = "Treatment") +
  geom_line(aes(color=factor(treated)), size=0.5) +
  scale_color_discrete(labels=c("Control", "Treated")) +
  geom_vline(xintercept = year_policy, lty=2) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

On average   

```{r}
mydata %>% 
  group_by(year, treated) %>% 
  summarise(y=mean(y),.groups="keep") %>% 
  ggplot(aes(x=year, y=y, group=treated, color = factor(treated))) + 
  annotate("rect", fill = "gray", alpha = 0.5,
           xmin = 2000, xmax = 2010,
           ymin = -Inf, ymax = Inf) +
  labs(title = paste("Outcome by year"),
       x = "Year", 
       y = "Outcome",
       colour = "Treatment") +
  geom_line() +
  scale_color_discrete(labels=c("Controls", "Treated")) +
  geom_vline(xintercept = year_policy, lty=2) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
```


