---
title: "An overview of Difference-in-Difference and Synthetic Control Methods: Classical and Novel Approaches"
description: |
  Society for Epidemiologic Research (SER) Workshop:   |  
  Staggered adoption: Part 4/4
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Roch Nianogo
    url: https://ph.ucla.edu/about/faculty-staff-directory/roch-nianogo
    affiliation: Department of Epidemiology, UCLA Fielding School of Public Health
    affiliation_url: https://ph.ucla.edu/about/faculty-staff-directory/roch-nianogo
date: "April 22th 2023"
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---


```{r setup, include=F}
knitr::opts_chunk$set(
	message = T,
	warning = T,
	#include = FALSE,
    dpi = 400,
  	echo = TRUE
)
if (!require("pacman")){
  install.packages("pacman", repos = 'http://cran.us.r-project.org')
} # a nice package to load several packages simultaneously


p_load("tidyverse","magrittr","broom",        #manipulate data
       "cleaR",                               #clearing workspace
       "here",                                #directory management
       "Synth", "gsynth",                     #synthetic control
       "panelView", "lme4", "estimatr",       #multi-level model
       "did",                                 #for staggered adoption
       "gtsummary")                           #for tables

p_load("flextable",                         # For tidying up regression output
        "broom.mixed",                      # For tidying up multilevel output
        "kableExtra",                       # For nice tables
        "knitr",                            # For nice tables
        "rmarkdown")   

p_load_gh("datalorax/equatiomatic")         # For nicely presenting regression coefficients

```

# Load data  

Let's load the data   

```{r load_data, warning=F, message=F}
mydata <- read_csv(here("data", "sim_data_hte_staggered.csv"))
```


In this new dataset, there are 50 states, 15 of which are treated and 35 untreated (controls)
The intervention was implemented at different times as below:  
 - state.name[1:5] ("Alabama", "Alaska", "Arizona", "Arkansas", "California") enacted policy in 2000  
 - state.name[6:10] ("Colorado", "Connecticut", "Delaware", "Florida", "Georgia") enacted policy in 2003  
 - state.name[6:10] ("Hawaii", "Idaho", "Illinois", "Indiana", "Iowa")  enacted policy in 2006


# Visualize the data  


```{r }
mydata <- read_csv(here("data", "sim_data_hte_staggered.csv"))

p_load("panelView")
panelview(y ~ treatedpost, data = mydata,
          index = c("state","year"), 
          pre.post = TRUE) 


panelview(y ~ treatedpost, data = mydata, 
          index = c("state","year"), 
          type = "outcome",  
          by.group = TRUE)


#For all the data
mydata %>% 
  ggplot(aes(x=year, y=y, group=state)) + 
  annotate("rect", fill = "gray", alpha = 0.5,
           xmin = 2000, xmax = Inf,
           ymin = -Inf, ymax = Inf) +
  labs(title = paste("Outcome by year"),
       x = "Year", 
       y = "Outcome",
       color = "Treatment") +
  geom_line(aes(color=factor(treated)), linewidth=0.5) +
  scale_color_discrete(labels=c("Control", "Treated")) +
  geom_vline(xintercept = 2000, lty=2) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 

```


# Analyze the data: Callaway and Sant’Anna   

For more information you can read the paper at [here](https://www.sciencedirect.com/science/article/pii/S0304407620303948?via%3Dihub)
The accompanying website also has a nice tutorial [here](https://cran.r-project.org/web/packages/did/vignettes/did-basics.html)


## load package and create new variables  

```{r callaway}

p_load(did)

mydata1 <- mydata %>% 
  mutate(first_treated = case_when( (state %in% state.name[1:5])   ~ 2000,
                                    (state %in% state.name[6:10])  ~ 2003,
                                    (state %in% state.name[11:15]) ~ 2006,
                                    TRUE~0))

paged_table(mydata1)
```

## Group-time effects  

```{r }
group_time_effects <- att_gt( yname  = "y",
                              tname  = "year",
                              idname = "state_num",
                              gname  = "first_treated",
                              xformla = ~ xit,
                              data = mydata1)



summary(group_time_effects)
ggdid(group_time_effects)
```

## Simple Aggregation  

```{r }
agg.simple <- aggte(group_time_effects, type = "simple")
summary(agg.simple)
```

## Dynamic Effects and (Event Studies): Effect by length of exposure  

```{r }
agg.es <- aggte(group_time_effects, type = "dynamic")
summary(agg.es)
ggdid(agg.es)
```


## Group-Specific Effects: Effect by group  

```{r }
agg.gs <- aggte(group_time_effects, type = "group")
summary(agg.gs)
ggdid(agg.gs)
# This is the group-specicf effects that can be used to estimate the overall ATT 
```

##  Calendar Time effects  

```{r }
agg.ct <- aggte(group_time_effects, type = "calendar")
summary(agg.ct)
ggdid(agg.ct)
```


# Analyze the data: Simple DID     

Now that we have seen different effects that could be obtained using the method 
above, let us see what other models give us   

```{r sdid}

p_load("estimatr")
dta <- lm_robust(y ~ treatedpost + factor(year) + xit, 
                 data = mydata,
                 fixed_effects=state,
                 clusters = state, 
                 se_type = "stata")
dta

did <- round(data.frame(ATT     = dta$coefficients["treatedpost"], 
                        se      = dta$std.error["treatedpost"],
                        low_ci  = dta$conf.low["treatedpost"],
                        high_ci = dta$conf.hig["treatedpost"]),2)
did


```
This quantity above is not giving us the group-specific overall ATT and could also
be biased. It is similar to the simple aggregation when using the `did` package

# Analyze the data: Generalized SCM  

```{r dscm}
y <- gsynth(y ~ treatedpost + xit, 
            data = mydata,  
            EM = F, 
            index = c("state","year"), 
            inference = "parametric", 
            se = TRUE,
            nboots = 100,  #so that it can run faster, default is 200
            r = c(0, 5), 
            CV = TRUE, 
            seed = 123,
            force = "two-way", 
            parallel = FALSE)

y1 <- round(data.frame(y$est.avg),2)


#Period-specific ATT
paged_table(y$est.att %>% as.data.frame())

#average ATT
y$est.avg


plot(y, type = "counterfactual", raw = "none", main="")

plot(y, type = "counterfactual", raw = "band", main="")
plot(y, type = "counterfactual", raw = "all")
plot(y, type = "ct", raw = "none", main = "", shade.post = FALSE)

```
This quantity above is not giving us the group-specific overall ATT. It is similar to the simple aggregation when using the `did` package. 
Another method is to obtain the overall ATT by estimating the effect in each unit (state) and then pool the estimates as done below  


# Analyze the data: Pooled effect  

## Method 1: Estimate single effects and pooled effects using the DID method
```{r did_pooled}

## Group 1: "Early adopters", policy enacted in 2000
data_filter_g1 <- mydata %>% 
  filter(state %in% state.name[1:5] | treated==0)

dta_g1 <- lm_robust(y ~ treatedpost + factor(year) + xit, 
                    data = data_filter_g1,
                    fixed_effects=state,
                    clusters = state, 
                    se_type = "stata")

did_g1 <- data.frame(group ="1-Early_adopters",
                     att     = dta_g1$coefficients["treatedpost"], 
                     se      = dta_g1$std.error["treatedpost"],
                     lowerci = dta_g1$conf.low["treatedpost"],
                     upperci = dta_g1$conf.hig["treatedpost"],
                     row.names = NULL)
did_g1


##Group 2: "Medium adopters", policy enacted in 2003

data_filter_g2 <- mydata %>% 
  filter(state %in% state.name[6:10] | treated==0)

dta_g2 <- lm_robust(y ~ treatedpost + factor(year) + xit, 
                    data = data_filter_g2,
                    fixed_effects=state,
                    clusters = state, 
                    se_type = "stata")

did_g2 <- data.frame(group = "2-Mid-adopters",
                     att     = dta_g2$coefficients["treatedpost"], 
                     se      = dta_g2$std.error["treatedpost"],
                     lowerci = dta_g2$conf.low["treatedpost"],
                     upperci = dta_g2$conf.hig["treatedpost"],
                     row.names = NULL)
did_g2


##Group 3: "ate adopters", policy enacted in 2006


data_filter_g3 <- mydata %>% 
  filter(state %in% state.name[11:15] | treated==0)

dta_g3 <- lm_robust(y ~ treatedpost + factor(year) + xit, 
                    data = data_filter_g3,
                    fixed_effects=state,
                    clusters = state, 
                    se_type = "stata")

did_g3 <- data.frame(group = "3-Late-adopters",
                     att    = dta_g3$coefficients["treatedpost"], 
                     se      = dta_g3$std.error["treatedpost"],
                     lowerci = dta_g3$conf.low["treatedpost"],
                     upperci = dta_g3$conf.hig["treatedpost"],
                     row.names = NULL)
did_g3

combined <- bind_rows(did_g1, did_g2, did_g3)
combined

p_load("metafor")
metaresult <- rma(yi = att, 
                  sei = se, 
                  data = combined, 
                  slab = group, 
                  method = "ML")

combined2 <- combined %>% 
  add_row(group="4-Overall",
          att=as.vector(metaresult$beta),
          se = metaresult$se,
          lowerci = metaresult$ci.lb,
          upperci = metaresult$ci.ub)

combined2
```


## Method 2: Estimate single effects and pooled effects using the generalized SCM

```{r gsynth_pooled}
#Create a function to estimate the effect of each state
gsynth_meta <- function(states, data, nboots = 200){
  
  
  dat <- data %>%
    filter(state=={{states}} | treated == 0)
  
  y <- gsynth(y ~ treatedpost + xit, 
              data = dat,  
              EM = F, 
              index = c("state","year"), 
              inference = "parametric", 
              se = TRUE,
              nboots = nboots, 
              r = c(0, 5), 
              CV = TRUE, 
              seed = 123,
              force = "two-way", 
              parallel = FALSE)
  
  y1 <- data.frame(y$est.avg)
  
  res <- tibble(
    state = states, 
    ATT=y1$Estimate,
    SE=y1$S.E.,
    lowerci=y1$CI.lower,
    upperci=y1$CI.upper)
  
  res
  
  return(res)
}

#test the function
gsynth_meta(states="California",
            nboots = 100,
            data=mydata)


#create the list and their intersection
states <- mydata %>% 
  filter(treated == 1) %>% 
  select(state) %>% 
  distinct() %>% 
  as.matrix() %>% 
  as.vector()
states


#create list of combinations
list_states <- expand_grid(states) %>% print(n=Inf)
list_states


#Loop through all the states
gsynth_res <- pmap_dfr(list_states, gsynth_meta, nboots=2, data =mydata)
#used nboots = 2 for simplicity. You should aim for at least 200 bootstrap samples
gsynth_res


#estimate the overall effect
metaresult2 <- rma(yi = ATT, 
                   sei = SE, 
                   data = gsynth_res, 
                   slab = state, 
                   method = "ML")


#combine the datasets
combined3 <- gsynth_res %>% 
  add_row(state="Overall",
          ATT=as.vector(metaresult2$beta),
          SE = metaresult2$se,
          lowerci = metaresult2$ci.lb,
          upperci = metaresult2$ci.ub)

combined3
#note that this has better standard errors compared to when pooling the 
#three groups

```


## Method 3: Estimate single effects and pooled effects using the Augmented Synthetic Control Methods  

```{r augsynth_methods}

p_load_gh("ebenmichael/augsynth")
set.seed(123)
augsynth.scm <-multisynth(y ~ treatedpost | xit,
                        unit = state, 
                        time = year, 
                        data = mydata, 
                        
                        fixedeff = T,
                        scm = T,
                        time_cohort = F, #can change this to T if interested in time cohorts instead of unit effects
                        progfunc="Ridge")
#progfunc = None for Traditional SCM,
#progfunc = Ridge for Ridge regression or augmented SCM
#progfunc = GSYN for the Generalized SCM

augsynth.scm 

res <- summary(augsynth.scm)
res

paged_table(res$att)
#is a dataframe that contains all of the point estimates, standard errors, and lower/upper confidence limits. Time = NA denotes the effect averaged across the post treatment periods.

plot(augsynth.scm)
plot(augsynth.scm, levels = "Average")
```